GPU-based Adaptive Beamforming for Active Sonar Imaging

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:08	00:38:47	(Tue 8.May)

However, despite its inherent potential, the \gls{MVDR} beamformer has yet to see widespread adoption in the active sonar community. There may be several reasons for this. For one, the method is not inherently robust, and will suffer from a phenomena called signal cancellation in active systems. Another reason is that in its original form, the computational complexity is cubic with the number of channels, O(M$^3$), while conventional beamformers are usually only at O($M$).

We will show that the suggested means for making the \gls{MVDR} behave well in active systems will also make it less computationally complex. Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single thread implementation in C.

@@

The fundamental task of sonar imaging is to estimate each pixel's individual contribution to the final recorded echo. If the data is digitally stored, we can do this by focusing on one pixel at a time in the post-processing. Since the geometric relationship of the array and image is known, one usually start by selecting delays that mitigates the difference in transmitter-pixel-receiver propagation time.

Consider an $M$ element phased array. Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights be  $w_m[n]$ and $\Delta_m[n]$. To simplify the upcoming notation, let sample $x_m[n]$ delayed by a factor $\Delta_m$ be written as:
\begin{align}
x^\Delta_m = x_m[n-\Delta_m[n]].
\end{align}
The beamformer output $z[n]$ is then simply the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x^\Delta_0[n]\\x^\Delta_1[n]\\\vdots\\x^\Delta_{M-1}[n]}.\label{z}
\end{align}
If the weights in (\ref{z}) were static, this would be referred to as the \gls{DAS} beamformer. The \gls{MVDR} beamformer~\cite{Capon1969}, however, finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction. The solution to this optimization problem is given as:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x[n]\H\}$ is the spatial covariance matrix. Since $\R$ is unknown, it must be estimated. We will do this by computing the sample covariance matrix, $\eR$. In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to achieve \gls{DAS}-like speckle statistics, and \emph{diagonal loading} to ensure a numerically well conditioned $\eR$~\cite{Synnevag2009a}.

\emph{Spatial averaging} is achieved by grouping array's channels into $M-L-1$ subarrays of length $L$, where all but one element overlap. If we let $x_l[n]$ represent the datavector from subarray $l$,
\begin{gather}
\x_l[n] = \bmat{x^\Delta_l[n] & x^\Delta_{l+1}[n] & \dots & x^\Delta_{l+L-1}[n]}\T,
\end{gather}
then the a ``subarray averaged'' sample covariance matrix, $\eR_\text{s}$, is given as:
\begin{gather}
\eR_\text{s}[n] = \frac{1}{M-L+1} \sumb{l=0}{M-L+1} \x_l[n]\x_l\H[n].\label{spatialR}
\end{gather}

@@

Because each pixel in the image can be processed independently, the \gls{MVDR} beamformer is well suited for parallel acceleration. A technology of particular interest here are \glspl{GPU}~\cite{Nilsen2009}, which tailored for image processing, available off-the-shelf at reasonable prices, and are found in most desktop computers already. The \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that executes a copy of a common function called a kernel.  However, there are some inherent drawbacks with such a design: Each core is left with a rather limited amount of local memory, and distributing data to the computing cores and collecting the result is a processes with some inherent latency.

@@

To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single thread C and Matlab implementation (\Fig{benchmarks}). Processing times for a 55\,kpixels sectorscan test image are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$. The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card. This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.

@@

As Figure \ref{benchmarks} illustrates, using a \gls{GPU} improves matters. If we look at the total runtimes for a subarray size of $L=16$ and $2N_\text{avg}+1=5$ pixels of temporal averaging, the \gls{GPU} implementation is able to process the test image in roughly 40\,ms. This translates to a processing capacity of 1.1\,Mpixels/s. For the same scenario the C implementation clocked in at roughly 3\,s and Matlab at 17\,s, which corresponds to a speedup factor of 75 and 435, respectively.

Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the \gls{GPU} implementation, and it gets even worse for larger $L$'s. However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky or UHDU solver should improve the runtimes further. Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curve of these implementation have a different curve than the \gls{GPU} implementation.

@@

The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}. Each pixel in the image can be processed independently, and the steps within each pixel can also

>>>>

However, despite its inherent potential, the \gls{MVDR} beamformer has yet to see widespread adoption in the active sonar community. There may be several reasons for this. For one, the method is not inherently robust, and will suffer from a phenomenon called signal cancellation in active systems. Another reason is that in its original form, the computational complexity is cubic with the number of channels, O(M$^3$), while conventional beamformers are at O($M$).

We will show that the suggested means for making the \gls{MVDR} behave well in active systems will also make it less computationally complex. Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single-thread implementation in C.

@@

The fundamental task of sonar imaging is to estimate each pixel's individual contribution to the final recorded echo. If the data is digitally stored, we can do this by focusing on one pixel at a time in the post-processing. Since the geometric relationship of the array and image is known, one usually start by selecting delays that compensates for the difference in transmitter-pixel-receiver propagation time.

Consider an $M$ element phased array. Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights be  $w_m[n]$ and $\Delta_m[n]$. To simplify the upcoming notation, let sample $x_m[n]$ delayed by $\Delta_m$ samples be written as:
\begin{align}
x^\Delta_m = x_m[n-\Delta_m[n]].
\end{align}
The beamformer output $z[n]$ is then simply the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x^\Delta_0[n]\\x^\Delta_1[n]\\\vdots\\x^\Delta_{M-1}[n]}.\label{z}
\end{align}
If the weights in (\ref{z}) were static, this would be referred to as the \gls{DAS} beamformer. The \gls{MVDR} beamformer~\cite{Capon1969}, however, finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction. The solution to this optimization problem is given as:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x[n]\H\}$ is the spatial covariance matrix. Since $\R$ is unknown, it must be estimated. We will do this by computing the sample covariance matrix, $\eR$. In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to achieve \gls{DAS}-like speckle statistics, and \emph{diagonal loading} to ensure a numerically well conditioned $\eR$~\cite{Synnevag2009a}.

\emph{Spatial averaging} is achieved by grouping array's channels into $M-L+1$ subarrays of length $L$, where all but one element overlap. If we let $x_l[n]$ represent the datavector from subarray $l$,
\begin{gather}
\x_l[n] = \bmat{x^\Delta_l[n] & x^\Delta_{l+1}[n] & \dots & x^\Delta_{l+L-1}[n]}\T,
\end{gather}
then the a ``subarray averaged'' sample covariance matrix, $\eR_\text{s}$, is given as:
\begin{gather}
\eR_\text{s}[n] = \frac{1}{M-L+1} \sumb{l=0}{M-L} \x_l[n]\x_l\H[n].\label{spatialR}
\end{gather}

@@

Because each pixel in the image can be processed independently, the \gls{MVDR} beamformer is well suited for parallel acceleration. A technology of particular interest here are \glspl{GPU}~\cite{Nilsen2009}, which are tailored for image processing, available off-the-shelf at reasonable prices, and are found in most desktop computers already. The \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that executes a copy of a common function called a kernel.  However, there are some inherent drawbacks with such a design: Each core is left with a rather limited amount of local memory, and distributing data to the computing cores and collecting the result is a process with some inherent latency.

@@

To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single-thread C and Matlab implementation (\Fig{benchmarks}). Processing times for a 55\,kpixels sectorscan test image are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$. The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card. This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.

@@

As Figure \ref{benchmarks} illustrates, using a \gls{GPU} improves matters. If we look at the total runtimes for a subarray size of $L=16$ and $2N_\text{avg}+1=5$ pixels of temporal averaging, the \gls{GPU} implementation is able to process the test image in roughly 40\,ms. This translates to a processing throughput of 1.1\,Mpixels/s. For the same scenario the C implementation clocked in at roughly 3\,s and Matlab at 17\,s, which corresponds to a speedup factor of 75 and 435, respectively.

Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the \gls{GPU} implementation, and it gets even worse for larger $L$'s. However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky or UHDU solver should improve the runtimes further. Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curve of these implementations have a different complexity curve than the \gls{GPU} implementation.

@@

The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}. Pixels in the image can be processed independently, and this also partially applies to the steps within each pixel.
